Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-07-18 paper/paper.Rmd
Derive volume density function as ratio

Now consider the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$ in the estimator.
For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ denotes a curve passing through $u$, and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$ for all $i=1,\dots,d$ and $\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}$. Then $\frac{Dw_1}{dt}(0)$ forms an orthonomal basis of the tangent space $T_pM$. 
The volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$ is defined as 
$$
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
$$

@@

Motivated by the estimator in \@ref(eq:denriem), we restate our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime H^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}^{-1}(\pmb{y}_i)$. Then the only problem left is the estimation of the Riemannian metric at each embedding point, discussed in \autoref{#riemmetric}.

[TODO: How to derive the volume density function?]

@@

@@

Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with raidus $r$. Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$, where $exp_p$ is the *exponential map* at point $p$. The exponential map maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{\prime}(0)=u$, and $\gamma(1)=exp_p(u)$. It is a differentiable bijective map of differentiable inverse (i.e. *diffeomorphism*). Intuitively, the exponential map moves point $p$ to an endpoint  at speed $u$ after covering the length of $\|u\|$ along the geodesic in one time unit.

>>>>

Now consider the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$ in the estimator.
For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ denote a curve passing through $u$, and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$ for all $i=1,\dots,d$ and $\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}$. Then $\frac{Dw_1}{dt}(0)$ forms an orthonomal basis of the tangent space $T_pM$. 
The volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$ is defined as 
\begin{equation}
\label{eq:volden}
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
\end{equation}

@@

Motivated by the estimator in \@ref(eq:denriem), we restate our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime H^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}^{-1}(\pmb{y}_i)$. Then the only problem left is the estimation of the Riemannian metric at each embedding point, discussed in \autoref{riemmetric}.

<!-- [TODO: How to derive the volume density function?] -->
By the definition of the volume density function in Equation \@ref(eq:volden), we could calculate how the volume change under the change of coordinates. When using the logarithm map, the coordinates of points on the manifold are transformed into their coordinates in the tangent space, i.e. geodesic normal coordinates. In this case, @Pelletier2005-vu gives the volume density function from the manifold to the tangent space which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, $|\det g(\exp^{-1}_p(q)) |$. 
In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding. Since the logarithm map of the manifold is also unknown, the volume density function in @Pelletier2005-vu is no longer valid. Instead, we need a volume density function independent of the chart.

For $p \in M$, denote an open ball centered at $p$ with radius $r$ as $B(p,r)$ and the open neighborhood of $0_p \in T_pM$ can be denoted as $B(0_p,r)=\exp_p^{-1}(B(p,r))$. Consider the Riemannian metrics $g^{\prime}$ and $g^{\prime \prime}$ in $B(0_p,r)$, with $g^{\prime} = \exp^{\ast}_p(g)$ being the pullback of $g$ induced by the exponential map and $g^{\prime \prime}$ being the canonical metric induced by $g_p$ in $B(0_p,r)$. 
Let $u \in B(0_p,r)$ be a tangent vector and for any given chart $(\overline{U}, \overline{\psi})$ of $B(0_p,r)$, we have $u \in \overline{U}$. 
The volumes of the parallelepiped spanned by the basis $\{(\partial / \partial\overline{\psi}_1|_u), \dots, (\partial / \partial\overline{\psi}_d|_u)\}$ are given by $| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime}$, and $| \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime \prime}$. The quotient of these two volumes is then independent of the chart $(\overline{U}, \overline{\psi})$ [@Henry2009-om]. 
Therefore, given a point $q \in B(p,r)$ and a tangent vector $u = \exp^{-1}_p(q) \in B(0_p,r)$, for any chart $(\overline{U}, \overline{\psi})$ of $B(0_p,r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
$$
\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} } \\ .
$$
The volume density function is only defined for points satisfying $d_g(p,q) < inj_gM$ so that there exists an open ball $B(p,r) \in T_pM$. And $\theta_p(q)=1$ for all $p,q \in \mathbb{R}^d$ when the manifold $M$ is $\mathbb{R}^d$ with the canonical metric.

Now we consider the case with the manifold learning embedding. Consider the chart $(U, \psi)$ induced by an orthonormal basis of the tangent space of the embedding $T_{\pmb{y}_i}E$ and a normal neighborhood of $\pmb{y}_i$. Given a query point $\pmb{p} \in B(\pmb{y}_i,r)$, the volume of the parallelepiped is spanned by the same basis.  
Then the volume density function is given by 
$$
\theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
$$ 
where $\pmb{H}$ is the estimated Riemannian metric for the unknown manifold.
Again, the volume density function is not defined for all $\pmb{p}$ but for points that satisfies $\pmb{p} \in B(\pmb{y}_i,r)$.

@@

As for the geodesic distance $d_g$,

@@

Denote $B(p, r) \subset M$ as an open ball centered at point $p$ with raidus $r$. Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$, where $exp_p$ is the *exponential map* at point $p$. The exponential map maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{\prime}(0)=u$, and $\gamma(1)=exp_p(u)$. It is a differentiable bijective map of differentiable inverse (i.e. *diffeomorphism*). Intuitively, the exponential map moves point $p$ to an endpoint  at speed $u$ after covering the length of $\|u\|$ along the geodesic in one time unit.
